A Journey from Wild to Textbook Data to Reproducibly Refresh the Wages Data from the National Longitudinal Survey of Youth Database

### 2022-05-11 paper/paper.Rmd
working on gender/sex explanation

According to @WickhamHadley2014TD, tidy data sets comply with three rules: (i) each variable forms a column, (ii) each observation forms a row, and (iii) each type of observational unit forms a table. The raw data, `new_data_qnames`, does not comply with these rules as it is organized such that each row corresponds to an individual. As respondents can have multiple jobs at specific years, the column names, such as `HRP1_1979`, `HRP2_1979`, `HRP1_1980` and `HRP2_1980`, contain the information about the job number up to 5 (`HRP1` = job 1, `HRP2` = job 2) and the year. The raw data consequently has a large number of columns (`r ncol(new_data_qnames)` to be specific). The values in the cell under the variables begin with `HRP` correspond to the hourly wage in dollars. A glimpse of this data shows:

@@

Thus, we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars, and the demographic variables. This is done using the `tidyverse` suite of packages [@tidyverse]: `tidyr` [@tidyr] to pivot the data into long-form, with `dplyr` [@dplyr] and `stringr` [@stringr] to create new variables from the downloaded data from the database, and code levels of factors by text wrangling. The long form of the data makes it possible to do these data transformations efficiently, and it is an intermediate step towards the final target data. The code for tidying the data are demonstrated at \url{`r yowie_pkgdown`/articles/raw-to-input-data.html} but also described in the subsequent subsections.

@@

In our final target data, we wish to include the demographic variables with variable names specified in brackets: sex (`sex`), race (`race`), age (`age_1979`), highest grade completed reported in each round of the survey (`grade`), highest grade completed ever reported (`hgc`), highest grade completed in terms of years, e.g., 9th grade = 9, 3rd-year college = 15, (`hgc_i`), highest grade completed in 1979 (`hgc_1979`) and whether the graduate equivalency diploma is obtained (`ged`).

For `sex` and `race`, we only rename the column names provided in the raw data, `new_data_qnames`. However, it is worth noting that the term `sex` is sometimes interchangeably used with the term `gender`. Therefore, when using the dataset in the classroom, the educators should emphasize this discourse that using this variables should be accompanied by special attention. Measuring gender as a binary variable has the potential to fail to capture people who do not identify themselves as either male or female or people whose gender does not align with their sex classification [@kennedy2020using]. From a statistical perspective, this sometimes difficult the adjustments of survey statistics to the population when gender is measured, for example, with three categories, and the census measured it with only two categories and used the term sex [@kennedy2020using]. Hence, for modern societal studies, it is highly recommended to separate sex and gender in the survey question and provide more options to respond to questions on gender.

>>>>

To get the data into tidy form [@WickhamHadley2014TD], it needs to comply with three rules: (i) each variable forms a column, (ii) each observation forms a row, and (iii) each type of observational unit forms a table. The raw data, `new_data_qnames`, violates these rules (i) and (ii) because information about an individual's multiple jobs over different years are in multiple columns. The raw data consequently has a large number of columns (`r ncol(new_data_qnames)` to be specific). The values in the cell under the variables begin with `HRP` correspond to the hourly wage in dollars. A glimpse of this data show the format problems:

@@

Thus, we re-arrange and wrangle the data into tidy data form, with columns corresponding to individual ID, year, job number, wage in dollars, and the demographic variables. This is done using the `tidyverse` suite of packages [@tidyverse]: `tidyr` [@tidyr] to pivot the data into long-form, with `dplyr` [@dplyr] and `stringr` [@stringr] to create new variables from the downloaded data from the database, and code levels of factors by text wrangling. The long form of the data makes it possible to do these data transformations efficiently, and it is an intermediate step towards the final target data. The code for tidying the data are demonstrated at \url{`r yowie_pkgdown`/articles/raw-to-input-data.html} but also described in the subsequent subsections.

@@

Our final target data will include the demographic variables (variable names): sex (`sex`), race (`race`), age (`age_1979`), highest grade completed reported in each round of the survey (`grade`), highest grade completed ever reported (`hgc`), highest grade completed in terms of years, e.g., 9th grade = 9, 3rd-year college = 15, (`hgc_i`), highest grade completed in 1979 (`hgc_1979`) and whether the graduate equivalency diploma is obtained (`ged`).

For `sex` and `race`, we have simplified the original names provided in the raw data, `SAMPLE_SEX_1979` and `SAMPLE_RACE_78SCRN`, respectively. However, it is important to note that the use of "sex" is probably not correct. This information is provided by the individual, and although only has two categories, is more consistent with "gender" as defined in  Therefore, when using the dataset in the classroom, the educators should emphasize this discourse that using this variables should be accompanied by special attention. Measuring gender as a binary variable has the potential to fail to capture people who do not identify themselves as either male or female or people whose gender does not align with their sex classification [@kennedy2020using]. From a statistical perspective, this sometimes difficult the adjustments of survey statistics to the population when gender is measured, for example, with three categories, and the census measured it with only two categories and used the term sex [@kennedy2020using]. Hence, for modern societal studies, it is highly recommended to separate sex and gender in the survey question and provide more options to respond to questions on gender.
